---
title: "Programming - Dates/Times and Functions-part 1"
output: html_notebook
---

#Ch. 16 Dates and Times
```{r}
library(tidyverse)

library(lubridate)
library(nycflights13)

today() # prints today's date as "2019-11-17"
now() #"2019-11-17 14:52:09 PST"

ymd("2017-01-31") #> [1] "2017-01-31"
mdy("January 31st, 2017") #> [1] "2017-01-31"
dmy("31-Jan-2017") #> [1] "2017-01-31"
ymd(20170131) #works without quotes
ymd_hms("2017-01-31 20:11:59") # the hms adds hours, minutes, seconds

flights %>% 
  select(year, month, day, hour, minute)
flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(departure = make_datetime(year, month, day, hour, minute))

make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>% 
  filter(!is.na(dep_time), !is.na(arr_time)) %>% 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>% 
  select(origin, dest, ends_with("delay"), ends_with("time"))

flights_dt

flights_dt %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 86400) # 86400 seconds = 1 day

flights_dt %>% 
  mutate(wday = wday(dep_time, label = TRUE)) %>% 
  ggplot(aes(x = wday)) +
    geom_bar()

flights_dt %>% 
  mutate(minute = minute(dep_time)) %>% 
  group_by(minute) %>% 
  summarise(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n = n()) %>% 
  ggplot(aes(minute, avg_delay)) +
    geom_line()
```
An alternative approach to plotting individual components is to round the date to a nearby unit of time, with floor_date(), round_date(), and ceiling_date()
```{r}
(datetime <- ymd_hms("2016-07-08 12:34:56")) #> [1] "2016-07-08 12:34:56 UTC"

year(datetime) <- 2020
datetime #> [1] "2020-07-08 12:34:56 UTC"
month(datetime) <- 01
datetime #> [1] "2020-01-08 12:34:56 UTC"
hour(datetime) <- hour(datetime) + 1
datetime #> [1] "2020-01-08 13:34:56 UTC"
```
If values are too big, they will roll-over

How does the distribution of flight times within a day change over the course of the year?
There is not much of a change.

Compare dep_time, sched_dep_time and dep_delay. Are they consistent? Explain your findings.
```{r}
flights_dt %>%
  mutate(dep_time_ = sched_dep_time + dep_delay * 60) %>%
  filter(dep_time_ != dep_time) %>%
  select(dep_time_, dep_time, sched_dep_time, dep_delay)
```
There are discrepancies. It looks like there are mistakes in the dates. The code would have had to check if the departure time is less than the scheduled departure time plus departure delay (in minutes). If they were consistent, then `dep_time = sched_dep_time + dep_delay`


Compare air_time with the duration between the departure and arrival. Explain your findings. (Hint: consider the location of the airport.)
```{r}
flights_dt %>%
  mutate(
    flight_duration = as.numeric(arr_time - dep_time),
    air_time_mins = air_time,
    diff = flight_duration - air_time_mins
  ) %>%
  select(origin, dest, flight_duration, air_time_mins, diff)
```


How does the average delay time change over the course of a day? Should you use dep_time or sched_dep_time? Why?
Use `sched_dep_time` because that is the relevant metric for someone scheduling a flight. Also, using `dep_time` will always bias delays to later in the day since delays will push flights later.


On what day of the week should you leave if you want to minimise the chance of a delay?
Saturday

What makes the distribution of diamonds$carat and flights$sched_dep_time similar?
```{r}
diamonds$carat 
flights$sched_dep_time
```
In both `carat` and `sched_dep_time` there are abnormally large numbers of values are at nice "human" numbers. In `sched_dep_time` it is at 00 and 30 minutes. In carats, it is at 0, 1/3, 1/2, 2/3,


Confirm my hypothesis that the early departures of flights in minutes 20-30 and 50-60 are caused by scheduled flights that leave early. Hint: create a binary variable that tells you whether or not a flight was delayed.
```{r}
flights_dt %>% 
  mutate(minute = minute(dep_time),
         early = dep_delay < 0) %>% 
  group_by(minute) %>% 
  summarise(
    early = mean(early, na.rm = TRUE),
    n = n()) %>% 
  ggplot(aes(minute, early)) +
    geom_line()
```
```{r}
# How old is Hadley?
h_age <- today() - ymd(19791014)
h_age

seconds(15) #> [1] "15S"
minutes(10) #> [1] "10M 0S"
hours(c(12, 24)) #> [1] "12H 0M 0S" "24H 0M 0S"
days(7) #> [1] "7d 0H 0M 0S"
months(1:6) #> [1] "1m 0d 0H 0M 0S" "2m 0d 0H 0M 0S" "3m 0d 0H 0M 0S" "4m 0d 0H 0M 0S"
#> [5] "5m 0d 0H 0M 0S" "6m 0d 0H 0M 0S"
weeks(3) #> [1] "21d 0H 0M 0S"
years(1) #> [1] "1y 0m 0d 0H 0M 0S"

years(1) / days(1) # estimate only: convert to intervals for accuracy [1] 365.25
```
Why is there months() but no dmonths()?
There is no direct unambiguous value of months in seconds

Explain days(overnight * 1) to someone who has just started learning R. How does it work?
The variable `overnight` is equal to `TRUE` or `FALSE`.
If it is an overnight flight, this becomes 1 day, and if not, then overnight = 0, and no days are added to the date.

Create a vector of dates giving the first day of every month in 2015. Create a vector of dates giving the first day of every month in the current year.
A vector of the first day of the month for every month in 2015:
```{r}
ymd("2015-01-01") + months(0:11)
floor_date(today(), unit = "year") + months(0:11)
```

Write a function that given your birthday (as a date), returns how old you are in years.
```{r}
e_age <- today() - ymd(19990613)
e_age #7462 days

age <- function(bday) {
  (bday %--% today()) %/% years(1)
}
age(ymd("1999-06-13")) # 20
```
Why can’t (today() %--% (today() + years(1)) / months(1) work?
The numerator of the expression is an interval with includes both a duration of time and a starting point. The denominator of the expression, `months(1)`, is a period, which is meaningful to humans but not defined in terms of an exact number of seconds.
```{r}
Sys.timezone() # [1] "America/Los_Angeles"
```
#Ch. 17 Intro to Pipes
#Ch. 16 Pipes!
```{r}
library(magrittr) # calls up pipe; it is included in the tidyverse
```
Pipes write like a story / like how we read them 

Ex. Little bunny Foo Foo
Went hopping through the forest
Scooping up the field mice
And bopping them on the head

foo_foo_1 <- hop(foo_foo, through = forest)
foo_foo_2 <- scoop(foo_foo_1, up = field_mice)
foo_foo_3 <- bop(foo_foo_2, on = head)

Helps avoid clutter & clear space
```{r}
diamonds <- ggplot2::diamonds
diamonds2 <- diamonds %>% 
  dplyr::mutate(price_per_carat = price / carat)

pryr::object_size(diamonds) #> 3.46 MB
pryr::object_size(diamonds2) #> 3.89 MB
pryr::object_size(diamonds, diamonds2) #> 3.89 MB
```
less typing = less mistakes

“tee” pipe %T>% works like %>% except that it returns the left-hand side instead of the right-hand side

```{r}
rnorm(100) %>%
  matrix(ncol = 2) %>%
  plot() %>%
  str()
#>  NULL

rnorm(100) %>%
  matrix(ncol = 2) %T>%
  plot() %>%
  str() # returns values
```

#Ch. 19 Functions
Writing a function has three big advantages over using copy-and-paste:

You can give a function an evocative name that makes your code easier to understand.

As requirements change, you only need to update code in one place, instead of many.

You eliminate the chance of making incidental mistakes when you copy and paste (i.e. updating a variable name in one place, but not in another).

Recommendation: write a function after writing a piece of code 3x

Ex.
```{r}
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
  
df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)
```

Why is TRUE not a parameter to rescale01()? What would happen if x contained a single missing value, and na.rm was FALSE?
The option `finite = TRUE` to `range()` will drop all non-finite elements, and `NA` is a non-finite element.
However, if both `finite = FALSE` and `na.rm = FALSE`, then this function will return a vector of `NA` values.

In the second variant of rescale01(), infinite values are left unchanged. Rewrite rescale01() so that -Inf is mapped to 0, and Inf is mapped to 1.
```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  y <- (x - rng[1]) / (rng[2] - rng[1])
  y[y == -Inf] <- 0
  y[y == Inf] <- 1
  y
}
rescale01(c(Inf, -Inf, 0:5, NA))
```

Practice turning the following code snippets into functions. Think about what each function does. What would you call it? How many arguments does it need? Can you rewrite it to be more expressive or less duplicative?

1) mean(is.na(x))
```{r}
mean_na <- function(x) {
  mean(is.na(x))
}
mean_na(c(0, 1, 2, NA, 4, NA))
```

2) x / sum(x, na.rm = TRUE)

sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)
Follow http://nicercode.github.io/intro/writing-functions.html to write your own functions to compute the variance and skew of a numeric vector.
```{r}
mean(is.na(x))
x / sum(x, na.rm = TRUE)
sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)
```
```

Write both_na(), a function that takes two vectors of the same length and returns the number of positions that have an NA in both vectors.
```{r}
both_na <- function(x, y) {
  sum(is.na(x) & is.na(y))
}
both_na(
  c(NA, NA, 1, 2),
  c(NA, 1, NA, 2)
)
both_na(
  c(NA, NA, 1, 2, NA, NA, 1),
  c(NA, 1, NA, 2, NA, NA, 1)
)
```


What do the following functions do? Why are they useful even though they are so short?

is_directory <- function(x) file.info(x)$isdir -> checks whether the path in `x` is a directory.
is_readable <- function(x) file.access(x, 4) == 0 -> checks whether the path in `x` is readable, meaning that the file exists and the user has permission to open it.
These functions are useful even though they are short because their names make it much clearer what the code is doing.

Read the complete lyrics to “Little Bunny Foo Foo”. There’s a lot of duplication in this song. Extend the initial piping example to recreate the complete song, and use functions to reduce the duplication.
foo_foo_1 <- hop(foo_foo, through = forest)
foo_foo_2 <- scoop(foo_foo_1, up = field_mice)
foo_foo_3 <- bop(foo_foo_2, on = head)
foo_foo_4 <- down(Good Fairy, speak = TRUE)
foo_foo_5 <- bunny(!see, foo_foo_3)

